Report

### 2022-07-15 manuscript.Rmd
start with moral politics data in tutorial

# Tutorial
This tutorial about Bayesian Evidence Synthesis (BES) requires installation of `r R.Version()$version.string`, Rstudio version `r RStudio.Version()$version` and package `bain` version `r packageVersion('bain')`. This tutorial focusses on synthesizing evidence for an informative hypotheses using the Product Bayes Factor (pbf). To illustrate the method, the `sesamesim` dataset will be used which is part of `bain`. For a detailed description of the dataset, run `?bain::sesamesim`.

@@

Let's say we are interested in whether the status of the site children grew up in (`site`) affects their score on a numbers test after watching Sesame Street for a year (`postnumb`). `postnumb` is a continuous variable which ranges from 0-63. `site` is a nominal variable indicating group membership with possible values 1 = disadvantaged inner city, 2 = advantaged suburban, 3 = advantaged rural, 4 = disadvantaged rural and 5 = disadvantaged Spanish speaking. In NHST we specify the null-hypothesis that the mean `postnumb` is equal for all groups in  site`:

@@

$H_i: \mu_{site2},\mu_{site3} > \mu_{site1},\mu_{site4},\mu_{site5}$

We can test the $H_0$ and $H_A$ simultaneously by running a linear regression on the data and passing the `lm` object with the specific hypotheses to bain.
```{r}
# load bain package
library(bain)

# convert `site` to nominal variable
sesamesim$site <- as.factor(sesamesim$site)

# specify model, allow meaningful zero-point
fit <- lm(postnumb ~ site-1, data = sesamesim)

#specify Hypotheses to evaluate
hypotheses <- "site2=site1=site3=site4=site5; site2&site3>site1&site4&site5"
hypotheses <- "site2=site1=site3=site4=site5"

@@

```
The output of `bain` seems to support neither hypothesis. The `fit` for both hypotheses is approximately zero which means that almost none of the data are in agreement with the restrictions the specified hypotheses pose. The `BF.c`, which is the Bayes Factor of the hypothesis versus its complement, is for both hypotheses close to zero. This implies that the complement of both hypotheses is $\frac{1}{BF.c}$ times more likely than the passed hypotheses. Finally, `PMPc`, the Posterior Model Probabilities of the specified hypotheses with $H_c$ included, is also close to zero for both specified hypothesis. This means that *out of* the specified hypotheses and their complement, the probability of $H_c$ being true is 99.9%.

### Single vs multiple studies.
Sometimes there is access to the data of not only a single, but multiple studies that all contain the necessary information to evaluate our hypothesis $H_i$. To evaluate $H_i$, it is possible to synthesize the support for $H_i$ over all studies using the PBF. In this example we split the `sesamesim` dataset up into three equal parts, to simulate three studies into the same topic.

@@

The function returns a `data.frame` containing the passed hypotheses in the rows and in the columns the PBF and the individual Bayes Factor for each sample individually. The `PBF` of both hypotheses is close to zero, meaning that the combined evidence for both hypotheses over all studies is low. When looking at the individual Bayes Factors, we also see that the PBF is lower than all of the Bayes Factors individually. For sample 3 the specified hypotheses are even more likely than their complement. This example illustrates another strength of the PBF; being able to evaluate a hypothesis more extremely than the most extreme individual evaluation. If all samples do not support a hypothesis, than the number of times the hypothesis is not supported should be included in its evaluation. This leads to a more extreme combined evaluation than any individual evaluation. Likewise, if there is very strong support in only a small minority of the samples, while the majority of samples shows no support at all, than it is likely that the hypothesis is not true.

@@

The `pbf()` function also accepts multiple `bain` objects.
This makes it possible to, for example, evaluate different sets of hypotheses for multiple data sets before using the resulting `bain` objects to aggregate the evidence for all hypotheses those datasets share in common.
The example below illustrates this princuple.
Each analysis has two hypotheses in common, but also contains one sample-specific hypothesis.
The `pbf()` function is called on the resulting `bain` object and returns output for the two common hypotheses.
Note that, in this case, `pbf()` does not require an argument `hypotheses`, as the hypotheses are contained in the `bain` objects.

```{r}
bain_sample1 <- bain(lm_sample1, "site2=site1=site3=site4=site5; site2>site1=site3=site4=site5; site2&site3>site1&site4&site5")
bain_sample2 <- bain(lm_sample2, "site2=site1=site3=site4=site5; site2=site1=site3>site4=site5; site2&site3>site1&site4&site5")
bain_sample3 <- bain(lm_sample3, "site2=site1=site3=site4=site5; site2=site1=site3=site4>site5; site2&site3>site1&site4&site5")

bain_allsamples <- list(bain_sample1, bain_sample2, bain_sample3)

pbf(bain_allsamples)

@@

```{r, echo = FALSE, eval = FALSE}
bains_different_hyps <- list(
  bain(lm_sample1, "site2=site1=site3=site4=site5"),
  bain(lm_sample2, "site2=site1=site3>site4=site5"),
  bain(lm_sample3, "site2&site3>site1&site4&site5")
)

@@

>>>>

# Tutorial
This tutorial about Bayesian Evidence Synthesis (BES) requires installation of `r R.Version()$version.string`, Rstudio version `r RStudio.Version()$version` and package `bain` version `r packageVersion('bain')`. This tutorial focusses on synthesizing evidence for an informative hypotheses using the Product Bayes Factor (pbf). To illustrate the method, a simulated counterpart of the data used in [REF Van Leeuwen et al. 2022] will be used avalable in `bain`. The data is used to investigate the relationship between political orientation and moral dispositions in three countries: the United states of America, Denmark and the Netherlands. A more detailed description of the dataset is either provided in [REF Van Leeuwen et al. 2022] or by running `?bain::synthetic_us`, `?bain::synthetic_dk` or `?bain::synthetic_nl`.

@@

Suppose we are interested in whether someone's socio-political orientation, i.e. liberal-conservative, is associated with the moral value of 'group loyalty'. Participants in all countries completed two scales that measure either socio-political orientation or group loyalty. Because each scale consists of multiple items of which some are reverse coded, single factors scores for both scales are first created for each country. Because the scales may differ per country, they also need to be standardized

```{r}
# reverse code
# create factor scores

# I create simple average scores for now
aggregate_scores <- function(country){
  df <- as.data.frame(cbind(
    scale(apply(country[,grepl("soc_", colnames(country))],1, mean)),
    apply(country[,grepl("grp_", colnames(country))],1, mean),
    apply(country[,grepl("her_", colnames(country))],1, mean)
    ))
  colnames(df) <- c('soc','grp', 'her')
  return(df)
  }

NLa <- aggregate_scores(NL)
USa <- aggregate_scores(US)
DKa <- aggregate_scores(DK)

Socio-political orientation and group loyalty have been aggregated to variables `soc` and `grp` respectively. Both are continuous variables with higher scores indicating conservatism or higher group loyalty respectively. In NHST we specify the null hypothesis that there is no association between `soc` and `grp`. We can test this hypothesis in each country by calculating a standardized regressoin coefficient.

$H_0: \beta_{grp} = 0$.

However, we may have prior expectations that higher group loyalty is somewhat related to  higher conservatism. An informative hypothesis could be:

$H_i: \beta_{grp} > .1$

<!-- SESAMESIM DATA: Let's say we are interested in whether the status of the site children grew up in (`site`) affects their score on a numbers test after watching Sesame Street for a year (`postnumb`). `postnumb` is a continuous variable which ranges from 0-63. `site` is a nominal variable indicating group membership with possible values 1 = disadvantaged inner city, 2 = advantaged suburban, 3 = advantaged rural, 4 = disadvantaged rural and 5 = disadvantaged Spanish speaking. In NHST we specify the null-hypothesis that the mean `postnumb` is equal for all groups in  site`:

@@

$H_i: \mu_{site2},\mu_{site3} > \mu_{site1},\mu_{site4},\mu_{site5}$ -->

Because it is very improbable that $\beta_{grp}$ is exactly equal to 0, more information can be won by testing $H_i$.
We can test $H_i$ by running a linear regression on the data and passing the `lm` object with the specific hypotheses to `bain`. For the Netherlands, the Bayes Factor would be calculated as follows:

```{r}
# load bain package
library(bain)

# specify model, allow meaningful zero-point
fit <- lm(soc ~ grp, data = NLa)

#specify Hypotheses to evaluate
hypotheses <- "grp > .1"

@@

```
<!-- SESAMESIM DATA: The output of `bain` seems to support neither hypothesis. The `fit` for both hypotheses is approximately zero which means that almost none of the data are in agreement with the restrictions the specified hypotheses pose. The `BF.c`, which is the Bayes Factor of the hypothesis versus its complement, is for both hypotheses close to zero. This implies that the complement of both hypotheses is $\frac{1}{BF.c}$ times more likely than the passed hypotheses. Finally, `PMPc`, the Posterior Model Probabilities of the specified hypotheses with $H_c$ included, is also close to zero for both specified hypothesis. This means that *out of* the specified hypotheses and their complement, the probability of $H_c$ being true is 99.9%. -->

### Single vs multiple studies.
<!-- SESAMESIM DATA: Sometimes there is access to the data of not only a single, but multiple studies that all contain the necessary information to evaluate our hypothesis $H_i$. To evaluate $H_i$, it is possible to synthesize the support for $H_i$ over all studies using the PBF. In this example we split the `sesamesim` dataset up into three equal parts, to simulate three studies into the same topic.

@@

The function returns a `data.frame` containing the passed hypotheses in the rows and in the columns the PBF and the individual Bayes Factor for each sample individually. The `PBF` of both hypotheses is close to zero, meaning that the combined evidence for both hypotheses over all studies is low. When looking at the individual Bayes Factors, we also see that the PBF is lower than all of the Bayes Factors individually. For sample 3 the specified hypotheses are even more likely than their complement. This example illustrates another strength of the PBF; being able to evaluate a hypothesis more extremely than the most extreme individual evaluation. If all samples do not support a hypothesis, than the number of times the hypothesis is not supported should be included in its evaluation. This leads to a more extreme combined evaluation than any individual evaluation. Likewise, if there is very strong support in only a small minority of the samples, while the majority of samples shows no support at all, than it is likely that the hypothesis is not true. -->

Sometimes there is access to the data of not only a single, but multiple studies that all contain the necessary information to evaluate our hypothesis $H_i$. To evaluate $H_i$, it is possible to synthesize the support for $H_i$ over all studies using the PBF calculated with `pbf()`. In this example, data to evaluate $H_i$ is available for three countries.

In the first use-case of `pbf()`, an `lm()` is ran for each country All `lm` objects are stored in a list which is passed to `pbf()`.
`pbf()` supports lists of all objects that `bain()` also supports. All supported types are specified in bain vignette *REF*. One exception is the use of named vectors, which will be covered later.
```{r}
fit_US <- lm(soc ~ grp + her, data = USa)
fit_DK <- lm(soc ~ grp + her, data = DKa)
fit_NL <- lm(soc ~ grp + her, data = NLa)

all_fits <- list(fit_US, fit_DK, fit_NL)

pbf(all_fits, hypotheses)
```

<!-- SESAME DATASET: The function returns a `data.frame` containing the passed hypotheses in the rows and in the columns the PBF and the individual Bayes Factor for each sample individually. The `PBF` of both hypotheses is close to zero, meaning that the combined evidence for both hypotheses over all studies is low. When looking at the individual Bayes Factors, we also see that the PBF is lower than all of the Bayes Factors individually. For sample 3 the specified hypotheses are even more likely than their complement. This example illustrates another strength of the PBF; being able to evaluate a hypothesis more extremely than the most extreme individual evaluation. If all samples do not support a hypothesis, than the number of times the hypothesis is not supported should be included in its evaluation. This leads to a more extreme combined evaluation than any individual evaluation. Likewise, if there is very strong support in only a small minority of the samples, while the majority of samples shows no support at all, than it is likely that the hypothesis is not true. -->

@@

The `pbf()` function also accepts multiple `bain` objects.
This makes it possible to, for example, evaluate different sets of hypotheses for multiple data sets before using the resulting `bain` objects to aggregate the evidence for all hypotheses those datasets share in common.
The example below illustrates this principle.
The analyses have one hypotheses in common ($H_i: \beta_{grp} > .1$), but also contains one sample-specific hypothesis regarding $\beta_{her}$.
The `pbf()` function is called on the list of  resulting `bain` objects and returns output for the common hypotheses.
Note that, in this case, `pbf()` does not require an argument `hypotheses`, as the hypotheses are contained in the `bain` objects.

```{r}
bain_US <- bain(fit_US, "grp > .1; her = 0")
bain_DK <- bain(fit_DK, "grp > .1; her > 0")
bain_NL <- bain(fit_NL, "grp > .1; her < 0")

bain_all <- list(bain_US, bain_DK, bain_NL)

pbf(bain_all)

@@

```{r, echo = FALSE, eval = FALSE}
bains_different_hyps <- list(
  bain(fit_US, "grp = 0"),
  bain(fit_DK, "grp > .1"),
  bain(fit_NL, "grp < .1")
)

@@

```{r}
library(bain)

### THESE DATA WILL BE USED (READ CASPARS WHATSAPP MESSAGES)
DK <- bain::synthetic_dk
US <- bain::synthetic_us
NL <- bain::synthetic_nl
?bain::synthetic_dk

# dat <- as.data.frame(cbind(
#   X = c(rnorm(1000,0,1), rnorm(1000, 0.2, 1)),
#   group = rep(0:1, each = 1000)
#   ))

# dat$group <- as.factor(dat$group)

# x <- dat[dat$group == 0, 1]
# y <- dat[dat$group == 1, 1]
# 
# ttest <- t_test(x,y, var.equal = TRUE, paired = FALSE)
# res <- bain(ttest, hypothesis = "y > x")
# 
# 
# fit_lm <- lm(X ~ group-1, data = dat)
# res_lm <- bain(fit_lm, hypothesis = "group1 > group0")
# 
# cor.test(y = dat$X, x = rep(0:1, each = 1000))
```
